---
title: "Exploring Socioeconomic and Demographic Trends in Armenia and Its Neighbors"
subtitle: "Data Visualization – Summer 2025 Project"
author: 
  - Artur Safrastyan
date: "04/08/2025"
output:
  pdf_document: default
editor_options: 
  markdown: 
    wrap: 72
documentclass: article
geometry: margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(ggplot2)
library(ggExtra)
library(reshape2)
library(GGally)
library(tidyr)
library(dplyr)
library(tidyverse)
library(forecast)
library(knitr)
library(ggplot2)
library(ggfortify)
library(dplyr)
library(fpp2)
library(shiny)
library(ggplot2)
library(DT)
```

```{r df, echo=FALSE}
df <- read.csv("data_r.csv")
```

```{r, echo=FALSE}

df_long_numeric <- df %>%
  select(year, where(is.numeric)) %>%
  pivot_longer(
    cols = -year,
    names_to = c("variable", "country"),
    names_pattern = "^(.*)_(arm|az|ge|est)$",
    values_to = "value"
  ) %>%
  pivot_wider(
    names_from = "variable",
    values_from = "value"
  )

df_long_events <- df %>%
  select(year, starts_with("notable_event")) %>%
  pivot_longer(
    cols = -year,
    names_to = "country",
    names_pattern = "^notable_event_(.*)$",
    values_to = "notable_event"
  )
df_long_categorical <- df %>%
  select(year, era,
         matches("^women_polit_participation_index_"),
         matches("^regime_type_")) %>%
  pivot_longer(
    cols = -c(year, era),
    names_to = c("variable", "country"),
    names_pattern = "^(.*)_(arm|az|ge|est)$",
    values_to = "cat_value"
  ) %>%
  pivot_wider(
    names_from = "variable",
    values_from = "cat_value"
  )
df_combined <- df_long_numeric %>%
  left_join(df_long_events, by = c("year", "country")) %>%
  left_join(df_long_categorical, by = c("year", "country"))
df_combined <- df_combined %>%
  mutate(country_full = recode(country,
                               "arm" = "Armenia",
                               "az"  = "Azerbaijan",
                               "ge"  = "Georgia",
                               "est" = "Estonia"))
```

## **Overview**

This report explores long-term socio-demographic, political, and public
health trends in Armenia and selected post-Soviet states from 1958 to
2022, with a focus on population, life expectancy, corruption,
urbanization, and gender-related indicators. It examines how Armenia,
despite facing war, blockade, and isolation, demonstrates demographic
resilience unusual for a state under chronic geopolitical stress. By
leveraging multiple datasets, the project contextualizes visual patterns
within broader political, historical, and economic frameworks. The
report aims to assess how well traditional metrics like birth rate, life
expectancy, and corruption indices capture the lived effects of regime
change, disaster, and modernization in Armenia.

Before delving into empirical findings, it is essential to understand
the structure and origin of the data used in this analysis. The dataset,
compiled from multiple international indicators, serves as the
foundation for evaluating Armenia’s long-term development trajectory.

## **Data description**

The dataset used in this project was manually compiled from multiple
sources hosted on the [Our World in
Data](https://ourworldindata.org/country/armenia) platform. It includes
annual time-series data for Armenia and several post-Soviet states
(e.g., Georgia, Azerbaijan, Estonia) from 1958 to 2022. Variables
encompass a wide range of demographic, socio-political, and economic
indicators such as:

-   Life Expectancy (total, male, female)

-   Birth and Death Rates

-   Child Mortality Rates

-   Sex Ratio at Birth

-   Urban Population Percentage Population

-   Totals Democracy and Corruption Indices

-   Regime Type

-   Notable Historical Events (coded manually)

Data was collected from over a dozen downloadable tables and merged
manually.

## **Literature review**

Numerous studies address the "transition crisis" in former Soviet
states, especially its impact on mortality, fertility, and governance.
Notably, Cornia and Paniccià (2000) describe the post-Soviet health
collapse as a result of rapid privatization and institutional voids.
Regarding corruption, Ledeneva (2006) explores the persistence of "blat"
(informal networks) in post-Soviet systems, complicating transparency
reforms. In terms of gender bias, Guilmoto (2009) , UNFPA Armenia (2013)
and Cambridge (2024) have reported Armenia’s skewed sex ratios and
cultural-son preferences. On urbanization reversals, Buckley and
Tsenkova (2001) note that post-socialist housing markets contributed to
unique migration trends not seen in Global North urbanism. Armenia’s
resilience, therefore, challenges conventional wisdom on the uniformity
of post-Soviet decline.

## Plots and Analysis

We begin our analysis by examining life expectancy trends. This
indicator reflects both health infrastructure and broader social
conditions, making it a valuable proxy for assessing national resilience

```{r, echo=FALSE}
df_combined <- df_combined %>%
  mutate(notable_event = case_when(
    country == "est" & year == 2004 ~ "eu_member",
    TRUE ~ notable_event
  ))

event_labels <- df_combined %>%
  filter(
    (country == "arm" & notable_event %in% c("karabakh_war1", "spitak_earth", "karabakh_war2")) |
    (country == "az"  & notable_event %in% c("karabakh_war1", "karabakh_war2")) |
    (country == "ge"  & notable_event == "civil_war/wars") |
    (country == "est" & notable_event == "eu_member")
  ) %>%
  distinct(year, total_life_expect, country_full, notable_event) %>%
  mutate(
    notable_event = recode(notable_event,
      "spitak_earth"     = "Spitak Earthquake", 
      "karabakh_war1"    = "End of The First\n Karabakh War", 
      "karabakh_war2"    = "Karabakh War II", 
      "civil_war/wars"   = "Georgian Civil War",
      "eu_member"        = "EU Membership"
    ),
    notable_event = factor(notable_event, levels = c(
      "Spitak Earthquake", 
      "End of The First\n Karabakh War", 
      "Karabakh War II", 
      "Georgian Civil War",
      "EU Membership"
    ))
  )

ggplot(df_combined, aes(year, total_life_expect, group = country)) +
  geom_vline(xintercept = 1991, linetype = "dashed", color = "seagreen") +
  annotate("text", x = 1991, 
           y = min(df_combined$total_life_expect, na.rm = TRUE) * 1.05, 
           label = "USSR collapse", vjust = 0.2, hjust = -0.04, size = 2.6, color = "seagreen") +

  geom_line(linewidth = 0.6) +

  geom_point(data = event_labels,
             aes(x = year, y = total_life_expect, color = notable_event),
             size = 2.5, inherit.aes = FALSE) +

  facet_wrap(~country_full) +
  labs(
  title = "Figure 1: Life Expectancy each year (1958–2022)",
  x = "Year",
  y = "Life Expectancy",
  color = "Notable Event"
) +
  theme_bw()


```

*Notable events shown include armed conflicts, political milestones, and
natural disasters. For a full list with casualty estimates and
references, see Appendix B.*

*To see a definition of Life Expectancy, see Appendix A.*

The above plot reveals life expectancy trends in four post-Soviet
republics-Armenia, Azerbaijan, Georgia, and Estonia—annotated with major
historical disruptions. All four countries exhibit a visible stagnation
or decline in life expectancy in the early 1990s, immediately following
the USSR’s dissolution in 1991. This decline, particularly sharp in
Armenia, Georgia, and Azerbaijan, aligns with what some economists and
demographers term the “transition crisis”—a collapse of public
institutions, health systems, and economic stability following the shift
to market economies. Estonia, in contrast, exhibits only a minor dip in
life expectancy. Estonia swiftly oriented itself toward EU integration,
adopting rapid liberalization and receiving Western institutional
support, culminating in EU accession in 2004, which is annotated on the
graph. Estonia's life expectancy trends suggest that Western
institutional alignment (gaining EU membership) had measurable impacts
on public health.\
Armed conflicts of the early 90s have had a major impact on the 3
countries of the Caucasus. The least impacted was Armenia with only a
minor dip and/or stagnation of life expectancy, in contrast, the same
war (the first Karabakh war) had a much stronger effect on Azerbaijan.
Georgia shows a prolonged stagnation in the 1990s, worse than Armenia or
Azerbaijan. The Georgian Civil War (1991–1993), along with conflicts in
Abkhazia and South Ossetia, and state collapse under Shevardnadze, led
to a near-total breakdown in social services.\
The most devastating occurrence in Armenia was the Spitak earthquake
which caused major death and destruction, but the sharp recovery the
following years shows that despite the scale of the tragedy, public
resilience and Soviet-era disaster response structures were temporarily
effective.\
Important to highlight the resilience of Armenia’s life expectancy as a
striking anomaly, especially when benchmarked against regional peers
like Georgia and even more developed post-Soviet states like Estonia.
This pattern defies expectations based on conventional demographic
predictors—armed conflict, economic blockade, and isolation typically
drive life expectancy downward. Yet Armenia exhibits a steady upward
trajectory, interrupted only by truly catastrophic events (Spitak
Earthquake in 1988 and Second Karabakh War in 2020).\
Some suggestions are listed below

-   Armenia inherited a high literacy rate, strong medical education,
    and preventive healthcare infrastructure from the USSR. Even though
    economic collapse degraded much of this, the cultural and
    institutional remnants persisted Armenia maintained universal basic
    education, widespread vaccination coverage, strong maternal and
    infant care culture.

-   Armenia benefits enormously from its global diaspora, which sends
    aid (up to 13% of GDP in some years), funds local clinics,
    hospitals, and medical aid, enables medical tourism. Diaspora aid
    partially offsets the blockade.

-   Armenians place high cultural value on education and family, which
    correlates with better maternal-child health, lower alcohol abuse,
    robust elderly care within extended families. There is evidence from
    global health that strong family systems mitigate the worst effects
    of poverty on life expectancy.

Armenia is one of the rare cases where war, blockade, and geopolitical
isolation have not translated into demographic collapse.\

The below plot shows the population change of 4 former Soviet countries
from 1958 to 2022.

**Armenia's Population Change:** Armenia's population peaked at nearly
3.7 million in the late 1980s but has declined sharply since
independence in 1991, to just over 3 million in 2021, largely as a
result of its decreasing fertility rate and negative net emigration
rate. The decline was driven by multiple factors: Many Armenian
emigrants in the early 1990s were people who had lost jobs after the
severe 1988 Spitak earthquake, refugees and people fleeing the economic
and energy crises of 1992-95. Almost half a million people left the
country during this period, and approximately 100,000 persons or 3
percent of the country's population, emigrated during the beginning of
1990s due to political internal and external instability. The continued
stagnation reflects ongoing emigration of working-age adults and
persistently low birth rates (shown later on).

```{r, fig.width=6, fig.height=4, fig.align = "center", echo = FALSE}
event_labels2 <- df_combined %>%
  filter(
    (country == "arm" & notable_event %in% c( "karabakh_war1")) |
    (country == "az"  & notable_event == "karabakh_war1") |
    (country == "ge"  & notable_event =="civil_war/wars")) %>%
  distinct(year, population, total_life_expect, country_full, notable_event) %>%
  mutate(
    population_millions = population / 1e6,
    notable_event = recode(notable_event,
      
      "karabakh_war1"   = "End of First \
      Karabakh War",
      "civil_war/wars"  = "Start of\
      Civil War",
      "russo_geo_war"   = "Russian Georgian War"
    )
  )


ggplot(df_combined, aes(year, population/1000000, group = country)) +
  geom_line(linewidth = 0.6) +
  facet_wrap(~country_full, scales = "free_y") +

  geom_vline(xintercept = 1991, linetype = "dashed", color = "red") +
  annotate("text", x = 1991, 
         y = min(df_combined$population, na.rm = TRUE) / 1e6 * 1.05, 
         label = "USSR collapse", angle = 0, vjust = 0.5, hjust = -0.04, size = 2.6) +
geom_point(data = event_labels2, 
           aes(x = year, y = population/1000000), 
           color = "blue", size = 2.5, shape = 18, inherit.aes = FALSE) +
  geom_text(data = event_labels2,
          aes(x = year,
              y = population/1000000,
              label = notable_event),
          size = 2.5, vjust = 2.1, hjust = 0.3, color = "blue", inherit.aes = FALSE) +

  labs(title = "Figure 2: Population by  year (1958-2022)",
       x = "Year", y = "Population in millions") +
  theme_bw()
```

**Azerbaijan's Population Change**: Azerbaijan's population growth
continued despite conflicts, likely because internal displacement from
Karabakh was absorbed within Azerbaijan's borders. Regarding population
counting, international organizations typically count territories based
on internationally recognized borders, so Karabakh's population would be
included in Azerbaijan's official statistics. Also, important to mention
the fact that Azerbaijan has limited and unreliable population
statistics due to media freedom being severely curtailed and official
numbers have been put under scrutiny, often being labeled as an
overestimation.

**Georgia's Population Change**: From 1992 the population began to
decline sharply due to civil war and economic crisis driven mass
migration throughout the 1990s and into the early 2000s, and Georgia
lost more than one million people or almost 20 per cent of its
population between independence and 2002. The emigration of ethnic
minorities has also left an impact. The number of Russians living in
Georgia has dropped significantly over the past 20 years. The 2002
census recorded nearly 68,000 ethnic Russians living in Georgia, a
figure that at the time constituted about 1.5 percent of the total
population. That total marked a steep decline from the 341,000 Russians
(6.3 percent of the population) counted in 1989. Like Armenia Georgia's
population has stagnated over the past decade.

**Estonia's Population Change**: Following independence, immigration
slowed greatly, and many Russians left the country, with a net
population loss of about 90,000 between 1989 and 1994, but the outflow
slowed in the late 1990s. Estonia also has a very low birth rate, a
problem crippling most EU members.

The below plot shows the difference between the sexes in Armenia from
1958-2022.

```{r, fig.width=5, fig.height=3, fig.align = "center", echo = FALSE}
library(dplyr)
library(ggplot2)


event_lines <- df %>%
  filter(notable_event_arm %in% c("spitak_earth", "karabakh_war2")) %>%
  distinct(year, notable_event_arm) %>%
  mutate(label = recode(notable_event_arm,
                        "spitak_earth" = "Spitak Earthquake",
                        "karabakh_war2" = "Karabakh War II"))


ggplot(df, aes(x = year)) +
  geom_vline(data = event_lines, aes(xintercept = year),
             linetype = "dashed", color = "seagreen4", linewidth = 0.5) +
  geom_vline(xintercept = 1991, linetype = "dashed", color = "purple") +

  geom_point(aes(y = male_life_expect_arm, color = "Male"),size=2) +
  geom_point(aes(y = female_life_expect_arm, color = "Female"),size=2) +
  
  


  annotate("text", x = 1991, 
           y = min(df_combined$male_life_expect, na.rm = TRUE) * 1.05, 
           label = "Spitak Earthquake", angle = 0, vjust = 0.2, hjust = 1.3, size = 2.3, colour = "seagreen") +
  annotate("text", x = 2020, 
           y = min(df_combined$male_life_expect, na.rm = TRUE) * 1.05, 
           label = "Karabakh War II", angle = 0, vjust = 0.2, hjust = 1.1, size = 2.3, colour = "seagreen")+
  annotate("text", x = 1991, 
           y = min(df_combined$total_life_expect, na.rm = TRUE) * 1.05, 
           label = "USSR collapse", angle = 0, vjust = 0.2, hjust = -0.04, size = 3,colour = "purple")+

  scale_color_manual(values = c("Male" = "steelblue2", "Female" = "red")) +

  labs(
    title = "Figure 3: Life Expectancy in Armenia by Gender (1958-2022)",
    x = "Year",
    y = "Life Expectancy",
    color = "Gender"
  ) +
  theme_bw() +
  theme(
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 8),
    plot.title = element_text(size=11)
  )


```

This gender-disaggregated data reveals nuances about how the Soviet
collapse differently impacted men and women in Armenia. The most
striking pattern is the stark divergence in recovery trajectories: while
female life expectancy maintained its steady upward trajectory with only
minor disruption around 1991, male life expectancy experienced a much
more severe and prolonged crisis. The gender gap in life expectancy
widened dramatically from about 7-8 years in the Soviet era to nearly 10
years by the early 2000s. This reflects the classic post-Soviet male
mortality crisis, though life expectancy declined dramatically in the
first half of the 1990s, particularly for males across the former USSR.
However, Armenia's pattern is notable because adult mortality patterns
in Armenia have been more favorable than in Russia, due to a large
extent to lower mortality from alcohol-related causes. The most
interesting aspect is the recent convergence trend, while male life
expectancy has recovered significantly since the early 2000s (except for
the 2020 war), female life expectancy has actually plateaued or slightly
declined in recent years, most likely due to the coronavirus pandemic.
This suggests that Armenia may have had improvements in male-specific
mortality causes (possibly violence, accidents, or cardiovascular
disease).

Urbanization reflects long-term economic and demographic shifts. In the
post-Soviet space, urban population percentages reveal both internal
migration patterns and the legacy of centrally planned development.

```{r, fig.width=6, fig.height=3, fig.align = "center",echo = FALSE}
df_long <- df %>%
  select(era, urban_population_perc_arm, urban_population_perc_az, urban_population_perc_ge) %>%
  pivot_longer(cols = starts_with("urban_population_perc_"),
               names_to = "country",
               names_prefix = "urban_population_perc_",
               values_to = "urban_pop")

df_long$era <- recode(df_long$era,
                      "post_soviet" = "Post Soviet",
                      "soviet" = "Soviet")

df_long$era <- factor(df_long$era, levels = c("Soviet", "Post Soviet"))

ggplot(df_long, aes(x = era, y = urban_pop)) +
  geom_boxplot(outlier.shape = NA) +
  stat_summary(fun = mean, geom = "point", shape = 20, color = "red", size = 2) +
  facet_wrap(~ country, labeller = as_labeller(c(arm = "Armenia", az = "Azerbaijan", ge = "Georgia"))) +
  labs(
    title = "Figure 4: Percentage of Population in Urban Areas Across Eras by Country (1958–2022)",
    x = "Era",
    y = "Population in Urban Areas (%)"
  ) +
  theme_bw() +
  theme(plot.title = element_text(size = 9))

```

The red dot represents the mean of the box plot of the respective era.

This urbanization plot reveals dramatically different Soviet
modernization trajectories and post-collapse urban strategies across the
three countries, reflecting their distinct economic structures and
political priorities.

**Armenia's Urbanization**: Armenia's Soviet urbanization shows the most
dramatic transformation - from 1929 to 1936, the government began the
process of industrialization, compelling many living in the rural
countryside to settle in urban areas. At the time of the republic's
dissolution, close to 70 percent of its population lived in urban areas.
For a post Soviet analysis see the below plot.\
**Azerbaijan's Urbanization:** Azerbaijan maintained relatively stable,
moderate urbanization levels throughout both periods. Azerbaijan's
population is approximately 57% urban and 43% rural currently, showing
continuity with Soviet patterns. This reflects Azerbaijan's oil-based
economy, which required less dramatic rural-to-urban population shifts
compared to Armenia's heavy industrialization strategy.

**Georgia's Urbanization**: Georgia shows the most dramatic increase in
urbanization - jumping from around 50% in the Soviet period to nearly
60% post-collapse. This might be explained by the large population drop
that Georgia has faced, especially in rural areas. This time-series plot
for Armenia provides crucial nuance that corrects and deepens our
earlier analysis of the country's urbanization patterns.

```{r,fig.width=5, fig.height=3,fig.align = "center",echo = FALSE}
ggplot(df, aes(x = year, y = urban_population_perc_arm)) +
  geom_line(color = "steelblue",linewidth = 0.7) +
  
  geom_vline(xintercept = 1991, linetype = "dashed", color = "red") +
  annotate("text", x = 1991, 
           y = min(df_combined$urban_population_perc, na.rm = TRUE) * 1.05, 
           label = "USSR collapse", angle = 0, vjust = 0.2, hjust = -0.1, size = 3, colour = "black") +

  geom_vline(xintercept = 1988, linetype = "dashed", color = "red") +
  annotate("text", x = 1988, 
           y = min(df_combined$urban_population_perc, na.rm = TRUE) * 1.05, 
           label = "Spitak Earthquake", angle = 0, vjust = 0.2, hjust = 1.05, size = 3, colour = "black") +

  labs(
    title = "Figure 5: Percentage of Population in Urban Areas in Armenia (1958–2022)",
    x = "Year",
    y = "Population in Urban Areas (%)"
  ) +
  
  theme_bw() + theme(plot.title = element_text(size=9))
```

*To see a definition of % of Urban Population, see Appendix A.*

The above time-series plot of Armenia's urbanization from 1958 to 2022
provides more nuance.\
The percentage of the population residing in urban areas peaked in 1988
with 68%. After the destructive earthquake, urbanization began to drop
up to the late 2010s.\
The drop in urbanization following 1988 was driven by a cascade of
crises:

-   1988 Spitak Earthquake destroyed 90% of buildings in Spitak, heavily
    damaged Gyumri and Vanadzor Armenia’s second and third largest
    cities.

    Killed \~25,000 people and displaced hundreds of thousands.

-   Collapse of the Soviet Union (1991) triggered mass
    deindustrialization; urban jobs disappeared overnight. Resulted in
    urban depopulation as survival in rural areas became more viable
    than in cash-starved cities.

-   First Nagorno-Karabakh War (1992–1994). Militarized much of the
    economy and society. Induced large-scale displacement of refugees,
    both ethnic Armenians from Azerbaijan and internally displaced
    people.

De-urbanization is rare globally. In most countries, urbanization is a
one-way path. Armenia is an exception due to the overlapping disasters
of earthquake, war, economic collapse, and emigration.

Beyond health and demographic outcomes, governance quality plays a
crucial role in development. The following plot explores changes in
Armenia’s political corruption levels and regime type over time, with
key events such as the Velvet Revolution marked for reference.

```{r, fig.align = "center", echo = FALSE}
ggplot(df, aes(year, political_corruption_index_arm)) +
  geom_vline(xintercept = 1991, linetype = "dashed", color = "red") +
  annotate(
    "text", x = 1991, 
    y = min(df$political_corruption_index_arm, na.rm = TRUE) * 1.05, 
    label = "USSR Collapse", angle = 0, vjust = 0.2, hjust = 1.1, size = 3, colour = "black"
  ) +
  geom_vline(xintercept = 2018, linetype = "dashed", color = "red") +
  geom_point(aes(color = regime_type_arm)) +
  annotate(
    "text", x = 2018, 
    y = min(df$political_corruption_index_arm, na.rm = TRUE) * 1.05, 
    label = "Velvet Revolution", angle = 0, vjust = 0.2, hjust = 1.05, size = 3, colour = "black"
  ) +
  labs(
    title = "Figure 6: Political Corruption Index in Armenia (1958–2022)",
    x = "Year",
    y = "Political Corruption Index",
    color = "Regime Type"
  ) +
  scale_color_discrete(labels = c(
    "Closed Autocracy", 
    "Electoral Autocracy", 
    "Electoral Democracy"
  )) +
  theme_bw() + theme(plot.title = element_text(size=11))

```

*To see a definition of Corruption Index and Regyme Type, see Appendix
A.*

The above plot shows the corruption and regime dynamics in Armenia
(1958–2022). At first glance, the relatively low corruption index during
the Soviet period (1958–1991) might appear paradoxical, given that
Armenia functioned under a closed autocratic regime. However, this
observation aligns with broader patterns observed in centralized
authoritarian systems, where the lack of transparency, limited press
freedom, and absence of civil society mechanisms result in significant
underreporting of corruption rather than its actual absence. Informal
practices such as nepotism, clientelism, and personal networks were
entrenched in Soviet governance structures, but rarely quantified or
acknowledged officially (Ledeneva, 2006). With the onset of Perestroika
and the political liberalization under Mikhail Gorbachev in the late
1980s, Armenia was nominally classified as an electoral democracy. This
classification persisted into the early post-independence years
following the USSR's dissolution in 1991. Nevertheless, by 1995, the
regime type was reclassified by international indices (e.g., V-Dem,
Freedom House) as an electoral autocracy, primarily due to repeated
allegations of electoral manipulation, constrained media, and the
consolidation of power among political elites. Correspondingly,
Armenia's political corruption index began to surge, peaking around
2001, and remaining persistently high—above 0.85—throughout the 2000s
and 2010s. A significant turning point occurred with the Velvet
Revolution in 2018. Following this, Armenia's regime classification
shifted back to electoral democracy, attributed to what international
observers described as the first genuinely competitive elections in the
country’s post-Soviet history. In parallel, the corruption index
underwent a sharp decline—from over 0.8 to approximately 0.3 by
2022—suggesting tangible reductions in elite corruption, increased
prosecutorial activity, and growing institutional responsiveness.

A more culturally sensitive indicator is the sex ratio at birth (SRB),
which may reveal subtle forms of gender bias. In Armenia, this metric
began to deviate from the biological norm in the 1990s, raising
questions about the role of policy, tradition, and post-conflict
militarization.

```{r,fig.width=5, fig.height=4, fig.align = "center",echo = FALSE}
ggplot(df, aes(x = year, y = sex_ratio_at_birth_arm)) +
  geom_line(color = "steelblue3", linewidth = 1) +
  

  geom_hline(yintercept = 105, linetype = "dashed", color = "black",linewidth=0.7) +
  annotate("text", x = 1965, y = 108,label = "Biological norm (105)", hjust = 0.2, vjust = 0.5, size = 3.5, color = "gray30") +

  geom_vline(xintercept = 1991, linetype = "dashed", color = "red") +
  annotate("text", x = 1992, y = 108, label = "Karabakh War I", angle = 0, vjust = 20,hjust=1.2, size = 2.5) +

  geom_vline(xintercept = 2016, linetype = "dashed", color = "red") +
  annotate("text", x = 2016, y = 108, label = "Reforms against \ 
           sex selective abortion", angle = 0, vjust = 7.5, hjust=1.05,size = 2.5) +


  labs(
    title = "Figure 7: Distortion in Armenia’s Sex Ratio at Birth (1958–2022)",
    x = "Year",
    y = "Male births for every 100 female birth"
  ) +
  scale_y_continuous(limits = c(90, NA)) +
  theme_bw() + theme(plot.title = element_text(size=11))
```

The plot above displays Armenia’s sex ratio at birth (SRB)—defined as
the number of male births per 100 female births—over time. While the
biologically normative range for SRB is approximately 105 males per 100
females, Armenia exhibits a systematic deviation from this natural
baseline, particularly after the early 1990s. The sharp rise in SRB
coincides with the aftermath of the First Nagorno-Karabakh War
(1988–1994) and Armenia’s transition to independence in 1991. This
distortion is attributed to sex-selective abortion, a phenomenon
observed in several post-Soviet and East Asian contexts. In Armenia’s
case, the drivers are multifaceted: Sociocultural pressures, including
strong son preference, rooted in patriarchal inheritance norms and
expectations of sons to provide for parents in old age. Militarized
nationalism, intensified by ongoing conflict with Azerbaijan, which
ascribes male offspring as future soldiers and protectors of the
homeland. The availability of prenatal sex-detection technologies,
coupled with limited legal or regulatory frameworks during the early
post-Soviet years, enabled the widespread use of sex-selective
termination.

Beginning in 2016, the Armenian government implemented a legal ban
specifically targeting sex-selective abortions, accompanied by public
awareness campaigns and health sector guidelines aimed at addressing
gender-biased reproductive choices. The data reflect a gradual
normalization of the SRB in the years following the intervention.
Nevertheless, the ratio remains elevated relative to the global norm,
indicating persistent cultural and structural biases.

The continued imbalance carries long-term demographic and social risks,
including gender imbalances in the adult population, skewed marriage
markets, and potential increases in gender-based violence. Moreover, the
systematic elimination of female fetuses, while not immediately apparent
in demographic totals, may gradually suppress Armenia’s population
growth in the coming decades, as the future base of mothers shrinks,
compounding the country’s existing challenges of low fertility and
emigration.

```{r, fig.width=6, fig.height=3,fig.align = "center", echo = FALSE}
df <- df %>%
  mutate(
    birth_rate_norm = (birth_rate_arm - min(birth_rate_arm, na.rm = TRUE)) /
                      (max(birth_rate_arm, na.rm = TRUE) - min(birth_rate_arm, na.rm = TRUE)),
    sex_ratio_norm = (sex_ratio_at_birth_arm - min(sex_ratio_at_birth_arm, na.rm = TRUE)) /
                     (max(sex_ratio_at_birth_arm, na.rm = TRUE) - min(sex_ratio_at_birth_arm, na.rm = TRUE))
  )

ggplot(df, aes(x = year)) +
  geom_line(aes(y = birth_rate_norm, color = "Birth Rate (0-1)"),linewidth=0.9) +
  geom_line(aes(y = sex_ratio_norm, color = "Male births
for every 100 female (0-1)"),linewidth=0.9) +
  labs(
    title = "Figure 8: Normalized Birth Rate vs. Sex Ratio at Birth (0–1 scale)",
    x = "Year",
    y = NULL,
    color = ""
  ) +
  theme_bw()+ theme(plot.title = element_text(size=9))

```

This plot presents the normalized trends of Armenia’s birth rate and sex
ratio at birth (SRB) from 1958 to 2022. A clear inverse relationship
emerges after the country’s independence: as birth rates declined
steadily, especially after the collapse of the USSR in 1991, the sex
ratio at birth began to deviate from the natural norm. This distortion
sharply increases in the mid-1990s and peaks around the early 2000s,
corresponding with widespread use of sex-selective abortions.

The plot proves that the abnormal rise of male births does not coincide
with a rise of birth rates, indicating an artificial factor's existence
as the cause of the problem.

## **Conclusion**

This report reveals Armenia’s demographic and governance trends to be
more resilient than commonly assumed. Despite experiencing one of the
longest-running blockades in modern history, two wars, a devastating
earthquake, and high emigration, Armenia’s life expectancy trends mirror
or surpass that of regional neighbors with ostensibly better
geopolitical positioning. The country’s inherited Soviet-era
infrastructure, cultural prioritization of education and healthcare, and
substantial diaspora support appear to have played key roles in
mitigating full collapse. However, challenges remain: declining birth
rates, persistent gender imbalances, and still high corruption rates
suggest ongoing structural vulnerabilities.

## **References**

-   Cornia, G. A., & Paniccià, R. (Eds.). (2000). *The mortality crisis
    in transitional economies*. Oxford University Press.

<!-- -->

-   Shkolnikov, V., Meslé, F., & Vallin, J. (1998). *Health crisis in
    Russia: recent trends in life expectancy and causes of death from
    1970 to 1993*. Population Studies.

<!-- -->

-   Ledeneva, A. V. (2006). *How Russia really works: The informal
    practices that shaped post-Soviet politics and business*. Cornell
    University Press.

<!-- -->

-   Guilmoto, C. Z. (2009). *The sex ratio transition in Asia*.
    Population and Development Review.

<!-- -->

-   UNFPA Armenia (2013). *Sex Imbalances at Birth in Armenia:
    Demographic Evidence and Analysis*.

<!-- -->

-   Cambridge (2024). *Isolating a culture of son preference among
    Armenian, Georgian and Azeri Parents in Soviet-era Russia*

## **Appendix**

**Appendix A** — Indicators and Definitions.

*Life Expectancy*: Average number of years a person is expected to live.

*Urbanization Rate*: % of population in areas defined as urban.

*Corruption index:* The Corruption Perceptions Index (CPI) is an index
that scores and ranks countries by their perceived levels of public
sector corruption, as assessed by experts and business executives. The
CPI generally defines corruption as an "abuse of entrusted power for
private gain".

*Regime Types:*

-   In *closed autocracies*, citizens do not have the right to either
    choose the chief executive of the government or the legislature
    through multi-party elections.
-   In *electoral autocracies*, citizens have the right to choose the
    chief executive and the legislature through multi-party elections;
    but they lack some freedoms, such as the freedoms of association or
    expression, that make the elections meaningful, free, and fair.
-   In *electoral democracies*, citizens have the right to participate
    in meaningful, free and fair, and multi-party elections.
-   In *liberal democracies*, citizens have further individual and
    minority rights, are equal before the law, and the actions of the
    executive are constrained by the legislative and the courts.

**Appendix B** — Major Events with Estimated Death Tolls and/or
Explanaitions

| Event | Year(s) | Estimated Deaths | Notes |
|----|----|----|----|
| Spitak Earthquake | 1988 | \~25,000 | Armenia’s deadliest natural disaster. Caused major death and infrastructural damage. |
| First Karabakh War | 1992-1994 | \~31,000. \~6000 from the Armenian side, \~25000 from the Azerbaijani side. | Ended in Armenian side's victory and control over Nagorno Karabakh. Had major demographic impact on Azerbaijan. |
| Georgian Civil War | 1991-1993 | \~20,000 | Resulted in near state collapse. Major impact on Georgian demographics. |
| Russo-Georgian War | 2008 | \~850 | Short but politically significant. |
| Karabakh War II | 2020 | \~7,000\* | Highly disruptive despite short duration. |

^\* Important to note Azerbaijan's official death toll is being heavily scrutinized and is considered as untrustworthy^
